---
title: "Interactive Visualisations"
subtitle: "Worksheet based on a graduate course on 'Electoral Behaviour' by Julia Schulte-Cloos"
format:
  html:
    theme: cosmo
    fontfamily: libertinus
    highlight-style: textmate
    smooth-scroll: false
    number-sections: false
    code-fold: true
    code-summary: "Show Code"
    collapsed: false
    toc: true
    toc-float: false
    standalone: true
    embed-resources: true
execute:
  warning: false
  message: false
  fig.showtext: true  
bibliography: "../literature.bib"
link-citations: true
---

```{r setup, include=FALSE}
pacman::p_load(tidyverse, 
               ggiraph, 
               gganimate, 
               dygraphs, 
               leaflet,
               showtext,
               sysfonts,
               gapminder,
              plotly)

# set global chunk options
knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 4.8,
  fig.align = "center",
  out.width = "90%",
  collapse = TRUE)

# specify cairo as a png device
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

# add a font from google

font_add_google("Gudea", "gudea")
showtext_auto()


theme_set(theme_light(base_size = 14, base_family = "gudea"))


```

## Interactive `ggplot` graphs with `ggiraph`

We start by generating an animated `ggplot2` that shows the exact value when hovering over a given point. The data is official election data, showing the average vote share of the German AfD during the 2019 EP elections by the federal states.

```{r ggiraph}

afdshare <- read_csv("./data/afdshare_by_states_ep19.csv")

afd_plot <- afdshare %>%
  ggplot(aes(y = state, x = mean_afd, color = state)) +
  geom_point(size = 2) +
  geom_pointrange(aes(
    color = state,
    xmin = mean_afd - (sd_afd),
    xmax = mean_afd + (sd_afd)
  )) +
  theme(
    panel.background = element_blank(),
    panel.grid.major = element_blank()
  ) +
  scale_colour_viridis_d() +
  labs(
    y = "",
    x = "Avg. AfD Share (EP 2019)"
  ) +
  theme(legend.position = "none") +
  geom_point_interactive(aes(tooltip = paste0(
    "<b>AfD Share: </b>",
    round(mean_afd, digits = 2)
  )),
  size = 1
  ) +
  theme(axis.text.y = element_text(margin = margin(t = 20, r = 20, b = 20, l = 20)))



tooltip_css <- "background-color:#f8f9f9;color:#1a1a1a;padding:10px;border-style:solid;border-width:2px;border-color:#125687;border-radius:5px;font-family: 'gudea', 'Signika', 'DIN', 'Helvetica', sans-serif;"

girafe(
  ggobj = afd_plot,
  options = list(
    opts_hover(css = "cursor:pointer;stroke:blue;fill-opacity:0.3"),
    opts_tooltip(css = tooltip_css, opacity = 0.9)
  ),
  width_svg = 6,
  height_svg = 4
)


```


## Interactive plots with `plotly`

`plot_ly` does not require a `ggplot` object as a starting point, but can be used to create interactive plots from scratch. Here, we use the built-in `USArrests` dataset to create a scatter plot showing the relationship between urban population and assault rates in different US states. 

```{r}

# Prepare the data
data <- USArrests %>%
  mutate(State = rownames(USArrests))

# Create the plot
plot_ly(data, 
        x = ~UrbanPop, 
        y = ~Assault,
        type = 'scatter',
        mode = 'markers',
        text = ~paste("State:", State,
                      "<br>Murder:", Murder,
                      "<br>Assault:", Assault,
                      "<br>Rape:", Rape),
        marker = list(
          size = ~Rape,
          color = ~Murder,
          colorscale = 'Viridis',
          showscale = TRUE,
          colorbar = list(title = "Murder Rate"),
          sizemode = 'diameter',
          sizeref = 2
        )
) %>%
  layout(
    title = "US Arrests by State (1973)",
    xaxis = list(title = "Urban Population (%)"),
    yaxis = list(title = "Assault Rate"),
    hovermode = "closest"
  )

```



## GIF-animations with `gganimate`

We load the data, which is a pre-processed dataset of more than 200.000 respondents who participated in the ESS survey between 2008 and 2018. The data contains information about the respondents' vote choice in the respective last national election and about their occupation, which was coded based on the Oesch class scheme. We only load the pre-processed data containing the average share of radical right supporters in each respective social class by ESS year.

```{r load-data}

ess_aggregated <- read_csv("./data/ess_radical_right_class_scheme.csv")

```

We then create a default `ggplot`, while adding the `gganimate` layer at the very end on top of this default `ggplot`.

```{r gganimate}

# create a named vector with the different classes and their colours that we want to use
animation_color_scale <- ess_aggregated %>%
  mutate(
    animation_color_scale =
      case_when(
        class_scheme == "Socio-Cultural Professionals" |
          class_scheme == "Production Workers" ~ "#0a1c70",
        TRUE ~ "#18a009"
      )
  ) %>%
  ungroup() %>%
  select(class_scheme, animation_color_scale) %>%
  dplyr::distinct() %>%
  pull(animation_color_scale, class_scheme)




animation_rrp <- ggplot(
  data = ess_aggregated,
  aes(
    y = class_scheme,
    x = rrp_vote,
    colour = class_scheme,
  )
) +
  geom_pointrange(aes(x = mean, xmin = conf_low, xmax = conf_high)) +
  scale_color_manual(
    values = animation_color_scale,
    guide = "none"
  ) +
  expand_limits(y = c(0, 0.20)) +
  scale_x_continuous("Average Share",
    breaks = c(seq(0, 0.20, 0.05))
  ) +
  labs(y = "") +
  # gganimate code
  transition_states(
    essyear,
    transition_length = 1,
    state_length = 0.5
  ) +
  enter_fade() +
  exit_shrink() +
  ease_aes("sine-in-out") +
  # gganimate specific title
  labs(title = "Radical right voters\nESS Year: {closest_state}") +
  theme(plot.margin = margin(1, 1, 3, 2, "cm"))

# show the animation
animation_rrp


```


## Timelines with `dygraphs`

We load a subset of a survey containing adolescents' attitudes towards the EU and their exact birthdates, which we show in a interactive `dygraph`. The [`dygraph` library](http://dygraphs.com/) is an open-source JavaScript charting library.

```{r load-birthdates-data}

birthdates_data <- read_csv2("./data/birthdates.csv")

dygraph(birthdates_data,
        main = "No of adolescents reaching voting age") %>%
  dyEvent("2004-06-15", "EP elections", labelLoc = "top") %>%
  dyOptions(fillGraph = TRUE, fillAlpha = 0.4)

```

We can also specify a range by relying on a pre-defined window:

```{r dygrpah-range}

dygraph(birthdates_data,
        main = "No of adolescents reaching voting age") %>%
  dyRangeSelector(dateWindow = c("1999-09-15", "2006-03-15")) %>%
  dyEvent("2004-06-15", "EP elections", labelLoc = "top") %>%
  dyOptions(fillGraph = TRUE, fillAlpha = 0.4)
```

## Interactive spatial visualisations with `leaflet`

The R package `leaflet` provides R bindings for the javascript mapping library leaflet.js.

```{r leaflet}

leaflet(height=500,
        width=850) %>% 
  addTiles() %>% 
  addMarkers(lat=47.4107931,
             lng=8.5402012,
             popup="UZH - IPZ") %>%
  setView(lat=47.4107931,
          lng=8.5402012, zoom=15)

```

We can also visualise multiple locations and data points. 

```{r}

# Create country coordinates dataframe
country_coords <- data.frame(
  country = c("Afghanistan", "Brazil", "China", "India", "United States", "Nigeria", "Germany", "Japan"),
  lat = c(33.93911, -14.2350, 35.8617, 20.5937, 37.0902, 9.0820, 51.1657, 36.2048),
  lon = c(67.70995, -51.9253, 104.1954, 78.9629, -95.7129, 8.6753, 10.4515, 138.2529)
)

# Filter Gapminder data for 2007 and join coordinates
gapminder_data <- gapminder %>%
  filter(year == 2007, country %in% country_coords$country) %>%
  left_join(country_coords, by = "country")

# Create leaflet map
leaflet(gapminder_data) %>%
  addTiles() %>%
  setView(lng = 0, lat = 20, zoom = 2) %>%
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    radius = ~sqrt(pop / 1000000),
    color = "blue",
    stroke = FALSE,
    fillOpacity = 0.5,
    popup = ~paste0("<strong>", country, "</strong><br>Population: ", format(pop, big.mark = ","))
  ) %>%
  leaflet::addLegend(
    position = "bottomright",
    pal = colorNumeric("Blues", domain = gapminder_data$pop),
    values = ~pop,
    title = "Population",
    labFormat = labelFormat(big.mark = ","),
    opacity = 1
  )
```


